---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Variable Length Markov Chains

<!-- badges: start -->
[![R-CMD-check](https://github.com/fabrice-rossi/mixvlmc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/fabrice-rossi/mixvlmc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`mixvlmc` implements variable length Markov chains (VLMC) and variable length Markov chains with covariates, as described in:

- [Bühlmann, P. and Wyner, A. J. (1999), Variable length Markov chains. Ann. Statist. 27 (2) 480-513](https://dx.doi.org/10.1214/aos/1018031204)
- [Zanin Zambom, A., Kim, S. and Lopes Garcia, N. (2022), Variable length Markov chain with exogenous covariates. J. Time Ser. Anal., 43 (2) 312-328](https://doi.org/10.1111/jtsa.12615)

## Installation
The package is only available from GitHub:
```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("fabrice-rossi/mixvlmc")
```

## Usage
### Variable length Marvkoc chains
Variable length Marvkov chains (VLMC) are sparse high order Markov chains. They can be used to model time series with discrete values (states) with a mix of small order dependencies for certain states and higher order dependencies for other states. For instance, with a binary time series, the probability of observing 1 a time $t$ could be constant whatever the past states if the last one (at time $t-1$) was 1, but could depend on states at time $t-3$ and $t-2$ if the state was 0 at time $t-1$. A collection of past states that determines completely the transition probabilities is a _context_ of the VLMC. Read `vignette("context-trees")` for details about contexts and context tree. 

VLMC with covariates are extension of VLMC in which transition probabilities (probabilities of the next state given the past) can be influenced by the past values of some covariates (in addition to the past values of the time series itself). Each context is associated to a logistic model that maps the (past values of the) covariates to transition probabilities.  

### Fitting a VLMC
The package is loaded in a standard way.
```{r}
library(mixvlmc)
## we use ggplot2 for graphics but this is not a dependency of the package
library(ggplot2)
```
The main function of VLMC is `vlmc()` which can be called on a time series represented by a numerical vector or a factor. 
```{r}
set.seed(0)
x <- sample(c(0L, 1L), 200, replace = TRUE)
model <- vlmc(x)
model
```
The default parameters of `vlmc()` will tend to produce overly complex VLMC in order to avoid missing potential structure in the time series. In the example above, we expect the optimal VLMC to be a constant distribution as the sample is independent and uniformly distributed (is has not temporal structure). The default paramaters give here an overly complex model, as illustrated by its text based representation
```{r}
draw(model)
```

The representation uses crude ascii art to display the contexts of the VLMC organized in to tree:

- the root `*` corresponds to an empty context;
- one can read contexts by following branches (represented by 2 dashes `--`) down to its end (the leaf): for instance $(1, 0, 1, 0, 0, 0)$ is one of the context of the tree. 

Here the context $(1, 0, 1, 0, 0, 0)$ is associated to the transition probabilities $(1, 0)$. This means that when one observe this context in the time series, it is _always_ followed by a 0. Notice that contexts are traditionally written from the most recent value to the oldest one. Thus, the context $(1, 0, 1, 0, 0, 0)$ corresponds to the sub time series $(0, 0, 0, 1, 0, 1)$.  

#### BIC based model selection

The VLMC above is obviously overfitting to the time series, as illustrated by the 0/1 transition probabilities. A classical way to select a good model is to maximize the [BIC](https://en.wikipedia.org/wiki/Bayesian_information_criterion). In `mixvlmc` this can be done easily by _pruning_ a complex VLMC using a combination of `cutoff` and `prune`, as follows:
```{r}
alphas <- cutoff(model)
pruned_models <- c(list(model), lapply(alphas, function(a) prune(model, a)))
bics <- sapply(pruned_models, BIC)
best_model <- pruned_models[[which.min(bics)]]
draw(best_model)
```

As expected, we end up with a constant model. 

In time series with actual temporal patterns, the optimal model will be more complex. As a very basic illustrative example, let us consider the `sunspot.year` time series and turn it into a binary one, with high activity associated to a number of sun spots larger than the median number. 
```{r}
sun_activity <- as.factor(ifelse(sunspot.year >= median(sunspot.year), "high", "low"))
```
We adjust a VLMC with a very conservative value of `alpha`:
```{r}
sun_model <- vlmc(sun_activity, alpha=0.1)
sun_model
```
Then we proceed as above:
```{r sunspots_bic}
sun_alphas <- cutoff(sun_model)
pruned_sun_models <- c(list(sun_model), lapply(sun_alphas, function(a) prune(sun_model, a)))
sun_bics <- sapply(pruned_sun_models, BIC)
sun_bic_fig <- data.frame(alpha=c(0.1, sun_alphas), BIC = sun_bics)
ggplot(sun_bic_fig, aes(x=alpha, y=BIC)) + 
  geom_line() 
```

The plot shows that simpler models are too simple as the BIC increases when pruning becomes strong enough. The best model remains rather complex (as expected based on the periodicity of the Solar cycle):
```{r}
best_sun_model <- pruned_sun_models[[which.min(sun_bics)]]
draw(best_sun_model)
```

### Fitting a VLMC with covariates
To illustrate the use of covariates, we use the power consumption data set included in the package. We consider a week of electricity usage as follows:
```{r active_power_week_5}
pc_week_5 <- powerconsumption[powerconsumption$week == 5, ]
elec <- pc_week_5$active_power
ggplot(pc_week_5, aes(x=date_time, y=active_power)) + 
  geom_line() + 
  xlab("t") +
  ylab("Activer power (kW)")
```
The time series displays some typical patterns of electricity usage:

- low active power at night (typically below 0.4 kW);
- standard use between 0.4 and 2 kW;
- peak use above 2 kW.

We build a discrete time series from those (somewhat arbitrary) thresholds:
```{r}
elec_dts <- cut(elec, breaks = c(0, 0.4, 2, 8), labels=c("low", "typical", "high"))
```

The best VLMC model is quite simple. It is almost a standard order one Markov chain, up to the order 2 context used when the active power is typical. 
```{r}
elec_vlmc <- vlmc(elec_dts)
e_alphas <- cutoff(elec_vlmc)
pruned_elec_vlmcs <- c(list(elec_vlmc), lapply(e_alphas, function(a) prune(elec_vlmc, a)))
e_bics <- sapply(pruned_elec_vlmcs, BIC)
best_elec_vlmc <- pruned_elec_vlmcs[[which.min(e_bics)]]
draw(best_elec_vlmc)
```

As pointed about above, low active power tend to correspond to night phase. We can include this information by introducing a day covariate as follows:
```{r}
elec_cov <- data.frame(day = (pc_week_5$hour >= 7 & pc_week_5$hour <= 17))
```
A VLMC with covariate is estimated using the `covlmc` function:
```{r}
elec_covlmc <- covlmc(elec_dts, elec_cov, min_size = 5, alpha=0.5)
draw(elec_covlmc, time_sep= "|")
```

Pruning a covlmc model is slightly more complicated than in the case of vlmc models. The main difficulty is that cut off values cannot be computed directly from a complex model as new values will appear during the pruning process.  A typical way to circumvent this difficulty is to recompute cut off values after each pruning operation, as follows:
```{r}
elec_covlmc_models <- list(elec_covlmc)
current_model <- elec_covlmc
alpha <- cutoff(current_model)[1]
while(TRUE) {
  new_model <- prune(current_model, alpha)
  elec_covlmc_models <- c(elec_covlmc_models, list(new_model))
  current_model <- new_model
  new_alphas <- cutoff(current_model)
  if(is.null(new_alphas) || all(new_alphas>=alpha)) {
    break
  } else {
    alpha <- max(new_alphas[new_alphas<alpha])
  }
}
elec_covlmc_bics <- sapply(elec_covlmc_models, BIC)
best_elec_covlmc <- elec_covlmc_models[[which.min(elec_covlmc_bics)]]
draw(best_elec_covlmc, model = "full", time_sep = " | ")
```

As in the VLMC case, the optimal model is simple. The two contexts `typical` and `high` do not use the covariate. The `low` context does not switch to a `high` context (hence the single row of parameters) but uses the covariate. As expected, the probability of switching from `low` to `typical` is larger during the day. 


