---
title: "Variable Length Markov Chains with Covariates (COVLMC)"
output: 
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Variable Length Markov Chains with Covariates (COVLMC)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mixvlmc)
library(ggplot2)
```

## Limitations of Variable Length Markov Chains

Variable Length Markov Chains (VLMC) are very useful to capture complex
structures in discrete time series as they can mix short memory with
long memory in a contextual way, leading to sparse models.

However, they cannot capture the influence of exogenous variables on the
behaviour of a time series. As a consequence, a VLMC adjusted on
sequences influenced by covariates could fail to capture interesting
patterns.

### A typical example

Let us focus again on the electrical usage data set used in the package
introduction, using the following two weeks of data:

```{r active_power_week_15_16}
pc_week_15_16 <- powerconsumption[powerconsumption$week %in% c(15, 16), ]
elec <- pc_week_15_16$active_power
ggplot(pc_week_15_16, aes(x = date_time, y = active_power)) +
  geom_line() +
  xlab("Date") +
  ylab("Activer power (kW)")
```

We build again a discrete time series using the following thresholds:

-   low active power at night (typically below 0.4 kW);
-   standard use between 0.4 and 2 kW;
-   peak use above 2 kW.

```{r}
elec_dts <- cut(elec, breaks = c(0, 0.4, 2, 8), labels = c("low", "typical", "high"))
```

Adjusting a VLMC to this sequence gives the following model (using BIC
for model selection):

```{r}
elec_vlmc_tune <- tune_vlmc(elec_dts)
best_elec_vlmc <- as_vlmc(elec_vlmc_tune)
draw(best_elec_vlmc)
```

Considering the apparent regularity of the time series, one may wonder
whether this model is really capturing the dynamics of power electricity
consumption. In particular the longest memory of 3 time steps, i.e., 30
minutes, may seem a bit short compared to the somewhat long periods of
stability.

## Theoretical aspects

VLMC with covariates have been introduced in [Variable length Markov
chain with exogenous covariates](https://doi.org/10.1111/jtsa.12615).
The core idea is to enable the conditional probabilities of the next
state given the context to depend on exogenous covariates.

### Variable memory

A COVLMC models a pair of sequences. The main sequence is denoted
$\mathbf{X}=X_1, X_2, \ldots, X_n, \ldots$. The random variables take
values in a finite state space $S$ exactly as in a standard VLMC. We
have *in addition* another sequence of random variables
$\mathbf{Y}=Y_1, Y_2, \ldots, Y_n, \ldots$ which take values in
$\mathbb{R}^p$. This latter assumption $Y_l\in \mathbb{R}^p$ is can be
relaxed to more general spaces.

A COVLMC puts restriction on the conditional probabilities of $X_n$
given the past of *both* sequences. To simplify the presentation, we
denote $X^k_l$ the sequence of random variables $$
X^k_m=(X_k, X_{k-1}, \ldots, X_m),
$$ and we use similar notations for the values taken by the variables
$x^k_m$, as well as for $Y^k_m$ and $y^k_m$. We use here the traditional
reverser ordering in which $X_k$ is the more recent variable and $X_m$
the oldest.

A pair of sequences $\mathbf{X}$ and $\mathbf{Y}$ is a COVLMC if there
is a maximal order $l_{\max}$ and a function $l$ from $S^{l_{\max}}$ to
$\{0,\ldots,l_{\max}\}$ such that for all $n>l_{\max}$ $$
\begin{multline}
\mathbb{P}(X_n=x_n\mid X^{n-1}_1=x^{n-1}_1, Y^{n-1}_1=y^{n-1}_1)=\\
\mathbb{P}\left(X_n=x_n\mid X^{n-1}_{n-l\left(x^{n-1}_{n-l_{\max}}\right)}=x^{n-1}_{n-l\left(x^{n-1}_{n-l_{\max}}\right)}, Y^{n-1}_{n-l\left(x^{n-1}_{n-l_{\max}}\right)}=y^{n-1}_{n-l\left(x^{n-1}_{n-l_{\max}}\right)}\right).
\end{multline}
$$ As in VLMC, the $\mathbf{X}$ process has a finite and variable
memory, but this memory applies to both $\mathbf{X}$ and $\mathbf{Y}$.
Notice that the memory order depends only on $\mathbf{X}$ and that no
assumptions are made on the temporal behaviour of $\mathbf{Y}$. Thus a
COVLMC on $\mathbf{X}$ and $\mathbf{Y}$ is not a VLMC on the pair
$(\mathbf{X}, \mathbf{Y})$ (which can make sense if $\mathbf{Y}$ is
discrete).

As for VLMC, the memory length function generates a *context* function
$c$ which keeps in the past the part needed to obtain the conditional
distribution: $c$ is a function from $S^{l_{\max}}$ to
$\bigcup_{k=0}^{l_{\max}}S^k$ given by $$
c(x^{n-1}_{n-l_{\max}})=x^{n-1}_{n-l\left(x^{n-1}_{n-l_{\max}}\right)}
$$ The image by $c$ of $S^{l_{\max}}$ is the set of contexts of the
COVLMC which is entirely specified by $l$ and one conditional
distribution by unique context.

As the notations are somewhat opaque at first, we can illustrate the
definition with a simple example. We consider a binary sequence (values
in $S=\{0, 1\}$) and a single numerical covariate $Y_t\in\mathbb{R}$. As
in the theoretical example in
`vignette("variable-length-markov-chains")` we assume that $l_{\max}=3$
and that $l$ is given by $$
\begin{align*}
l(0, a, b)&=1&\forall a, \forall b,\\
l(1, 1, c)&=2&\forall c,\\
l(1, 0, 0)&=3,&\\
l(1, 0, 1)&=3.&\\
\end{align*}
$$ In practice the COVLMC is therefore fully described by specifying the
following probabilities: $$
\begin{align*}
\mathbb{P}(X_t=1\mid& X_{t-1}=0, Y_{t-1}=y_{t-1})\\
\mathbb{P}(X_t=1\mid& X_{t-1}=1, X_{t-2}=1, Y_{t-1}=y_{t-1}, Y_{t-2}=y_{t-2})\\
\mathbb{P}(X_t=1\mid& X_{t-1}=1, X_{t-2}=0, X_{t-3}=0, Y_{t-1}=y_{t-1}, Y_{t-2}=y_{t-2}, Y_{t-3}=y_{t-3})\\
\mathbb{P}(X_t=1\mid& X_{t-1}=1, X_{t-2}=0, X_{t-3}=1, Y_{t-1}=y_{t-1}, Y_{t-2}=y_{t-2}, Y_{t-3}=y_{t-3})
\end{align*}
$$

### Conditional distributions

The main difficulty induced by COVLMC compared to VLMC is the
specification of the conditional distributions. Indeed the conditional
distributions depend on $\mathbf{Y}$ and cannot simply be given by a
table. For instance, in the example above, we need to specify, among
others, $\mathbb{P}(X_t=1\mid X_{t-1}=0, Y_{t_1}=y_{t-1})$, for each all
values of $y_{t-1}\in \mathbb{R}$ (in $\mathbb{R}^p$ in the general
case).

A natural choice in this particular example is to use a logistic model,
i.e. to assume $$
\mathbb{P}(X_t=1\mid X_{t-1}=0, Y_{t_1}=y_{t-1})=g(\alpha^0+y_{t-1}\beta_1^0),
$$ with $g(t)=\frac{1}{1+\exp(-t)}$ is the logistic function. The
superscript on $\alpha^0$ and $\beta^0_1$ refer to the context, here
$0$, while the subscript on $\beta^0_1$ refers to the time delay (here
$1$). By extension, we have for example $$
\begin{multline*}
\mathbb{P}(X_t=1\mid X_{t-1}=1, X_{t-2}=1, Y_{t_1}=y_{t-1}, Y_{t_2}=y_{t-2})=\\
g\left(\alpha^{1,1}+y_{t-1}\beta^{1,1}_1+y_{t-2}\beta^{1,1}_2\right).
\end{multline*}
$$

More generally, the probability distribution associated to a context
could be given by any function that maps the values of the covariates to
a distribution on $S$, the state space. Following the original
[paper](https://doi.org/10.1111/jtsa.12615), `mixvlmc` uses multinomial
logistic regression as implemented by `VGAM::vglm()` or
`nnet::multinom()`, or a logistic regression provided by `stats::glm()`
for state spaces with only 2 states. This has several advantages over a
more general solution:

1)  during the estimation phase, one probability distribution is
    estimated for each relevant context: this could induce a large
    computational burden for more complex models than multinomial
    logistic ones;
2)  having a simple model enables to fit contexts with limited number of
    occurrences which is important to allow searching for long term
    dependencies;
3)  logistic models with different memory order are easy to compare
    using a likelihood-ratio test.

The last point is used in `mixvlmc` (as in the original paper) to
simplify local models with respect to the covariates. For a context of
length $l$, the probability distribution is assumed to depend on the $l$
past values of $\mathbf{Y}$ but in practice we allow a dependency to
only $k<l$ past values. For instance, we could have $$
\mathbb{P}(X_t=1\mid X_{t-1}=1, X_{t-2}=1, Y_{t_1}=y_{t-1}, Y_{t_2}=y_{t-2})=g\left(\alpha^{1,1}+y_{t-1}\beta^{1,1}_1\right).
$$

### Beta-Context algorithm

Estimating a COVLMC model from two time series is more complex than
estimating a VLMC model. `mixvlmc` implements the Beta-Context algorithm
proposed in the original [paper](https://doi.org/10.1111/jtsa.12615). It
is inspired from the context algorithm used for VLMC (and proposed in
[Variable length Markov
chains](https://dx.doi.org/10.1214/aos/1018031204)). It can be
summarized as follows:

1)  the first step consists in building a context tree (see
    `vignette("context-trees")`) from the $\mathbf{X}$ discrete
    sequences, almost exactly as for a VLMC: the only difference is that
    to be kept in the tree a context must appear a number of times that
    depends on its length, on the dimension of the covariates and on the
    number of states. This guarantees a minimal number of observations
    for the (maximum likelihood) estimation of the context dependant
    multinomial logistic regression.
2)  the second step is an estimation one: a multinomial logistic
    regression model is estimated for each context, using a number of
    past values of $\mathbf{Y}$ equal to the length of the context.
3)  the rest of the algorithm consists in pruning the context tree and
    the logistic models:
    1)  leaf contexts are first assessed in terms of model
        simplification. The likelihood-ratio test is used to decide
        whether the oldest value of $\mathbf{Y}$ is relevant or not.
        Essentially we compare e.g.
        $g\left(\alpha^{1,1}+y_{t-1}\beta^{1,1}_1\right)$ to
        $g\left(\alpha^{1,1}+y_{t-1}\beta^{1,1}_1+y_{t-2}\beta^{1,1}_2\right)$
        as an estimator of
        $\mathbb{P}(X_t=1\mid X_{t-1}=1, X_{t-2}=1, Y_{t_1}=y_{t-1}, Y_{t_2}=y_{t-2})$;
    2)  based on the results of those tests, it may be possible to
        completely remove a context from the tree, see the
        [paper](https://doi.org/10.1111/jtsa.12615) for details.

This last pruning phase is carried out repeatedly as context removals
turn internal contexts into leaves that can then be further simplified.

## COVLMC in practice

### Estimation

COVLMC estimation is provided by the `covlmc()` function. We build first
a simple example data set as follows:

```{r}
set.seed(0)
nb_obs <- 200
covariates <- data.frame(y = runif(nb_obs))
x <- 0
for (k in 2:nb_obs) {
  ## we induce a simple dependency to the covariate
  ## and an order 1 memory
  if (covariates$y[k - 1] < 0.5) {
    if (x[k - 1] == 0) {
      x[k] <- sample(c(0, 1), 1, prob = c(0.7, 0.3))
    } else {
      x[k] <- sample(c(0, 1), 1, prob = c(0.3, 0.7))
    }
  } else {
    if (x[k - 1] == 0) {
      x[k] <- sample(c(0, 1), 1, prob = c(0.1, 0.9))
    } else {
      x[k] <- sample(c(0, 1), 1, prob = c(0.5, 0.5))
    }
  }
}
```

Then we estimate a COVLMC as follows:

```{r}
model <- covlmc(x, covariates)
model
```

The estimation process is controlled by three parameters:

-   `max_depth`: the largest order/memory considered for the COVLMC
    (defaults to 100). This parameter is essentially a computational
    burden control parameter. It does not play a major role in COVLMC
    estimation because of the constraints imposed by `min_size`. The
    default value is very conservative;
-   `min_size`: this parameter controls the minimal number of
    occurrences needed for a context to be included in the initial
    context tree. It gives the number of occurrences *per parameter* of
    the logistic model which depends on both the length of the context,
    the dimension of the covariates and the number of states;
-   `alpha`: is the parameter of the pruning process of the Beta-Context
    algorithm. Pruning decisions are all based on likelihood ratio tests
    and `alpha` is the common level of all those tests.

The default parameters work well on the previous example, as shown by the obtained
model:
```{r}
draw(model, model = "full", p_value = FALSE)
```

The model has 2 contexts, 0 and 1, as expected. The logistic models are described
by their parameters. For context 0, the intercept is negative and the coefficient of
$y_{t-1}$ is positive: the probability of switching to 1 is small when $y_{t-1}$ is
small and increases with $y_{t-1}$. For context 1, the situation is reversed and
the effect of $y_{t-1}$ is smaller. This is consistent with the way the series
were constructed. 

Notice however that obtained an interesting model with the default parameters 
should not be seen as a general property and proper model choice must be implemented. 

### Model choice
As COVLMC estimation fits a potentially large number of logistic models to the 
data, the use of a penalized likelihood approach is recommended to set its parameters
and avoid overfitting. 

However, model choice is more complex in the case of the COVLMC than for VLMC. In particular,
the pair `cutoff()`/`prune()` does not work as well for COVLMC than for VLMC (see `vignette("variable-length-markov-chains")`). Indeed the pruning process of the 
Beta-Context algorithm is such that its effects cannot be predicted as easily as 
the ones of the Context algorithm. In practice, computing the largest `alpha` 
(test level) that is guaranteed to make a minimal but actual pruning of a given
COVLMC is easy. But subsequent cut off values could be misleading. To avoid any 
problem it is therefore recommended to rely on the `tune_covlmc()` function that
has been designed to explore the full "pruning space" associated to a given data set.

Used on the artificial example above, it gives:
```{r}
model_tune <- tune_covlmc(x, covariates)
model_tune
```
and
```{r}
draw(as_covlmc(model_tune), model = "full", p_value = FALSE)
```
The resulting model is the same as the one obtained before but it was properly 
obtained by minimising the BIC.

As for `tune_vlmc()`, the `max_depth` parameter is automatically increased to
avoid using it inadvertently as a regularisation parameter. However, while 
`min_size` is generally considered as not having a major influence on the model
selection for VLMC, this is not the case for COVLMC. There is currently no
support in `mixvlmc` for an automatic choice of `min_size` and one should therefore
test several values and compare the obtained BIC/AIC. In the articial case, we
can try for instance `min_size=2` as follows:
```{r}
model_tune_2 <- tune_covlmc(x, covariates, min_size = 2)
model_tune_2
```
and `min_size=10` as follows:
```{r}
model_tune_10 <- tune_covlmc(x, covariates, min_size = 10)
model_tune_10
```

Here we see no particular effect of the parameter because of the simplicity of the
problem.

Let us now come back to the electricity consumption example described above. We
introduce a very basic day/night covariate as follows:
```{r}
elec_cov <- data.frame(day = (pc_week_15_16$hour >= 7 & pc_week_15_16$hour <= 18))
```
and fit a COVLMC with
```{r}
elec_tune <- tune_covlmc(elec_dts, elec_cov)
elec_tune
```

The model selection process can be represented graphically as follows:
```{r}
ggplot(elec_tune$results, aes(x = alpha, y = BIC)) +
  geom_line() +
  geom_point()
```

The final model is:
```{r}
draw(as_covlmc(elec_tune), model = "full", p_value = FALSE, with_state = TRUE)
```

It shows some interesting patterns:

- one of the logistic model is degenerate: in the `high` context, no transition 
  to a `low` context can happen. This was already observed with the VLMC;
- the only dependency with respect to the covariate is in the context typical, 
  typical, typical. In this case, the probability to stay in the typical context is 
  increased during day time and decreased (but to a lesser extent) during the night. 
  In practice, this means that the model is able to generate longer sequences that
  stay in the typical state during the day than at night.

Notice finally that for this real world example, the `min_size` parameter has
some influence on the results. Setting it to a smaller value does not change the 
final model, as shown below:

```{r}
elec_tune_3 <- tune_covlmc(elec_dts, elec_cov, min_size = 3)
elec_tune_3
```

However, increasing the parameter to, e.g., 10 generates a simpler but weaker model
as show here:

```{r}
elec_tune_10 <- tune_covlmc(elec_dts, elec_cov, min_size = 10)
elec_tune_10
```


### Diagnostics
The package provides numerous ways to analyse a COVLMC. asic functions include

- `states()` returns the state space of the model;
- `depth()` returns the length of the longest context in the model;
- `context_number()` returns the number of contexts in the model.

For instance, the large model obtained above has the following
characteristics:
```{r}
elec_model <- as_covlmc(elec_tune)
states(elec_model)
depth(elec_model)
context_number(elec_model)
```
VLMC objects support classical statistical functions such as:
```{r}
logLik(elec_model)
AIC(elec_model)
BIC(elec_model)
```

### Contexts
The model can be explored in details by drawing its context tree 
(see `vignette("context-trees")` for details) as follows:
```{r}
draw(elec_model)
```

The `draw.covlmc()` function is more advanced than its VLMC counterpart. It provides more detailed information, particularly regarding p-values associated with pruning operations. The "merging" p-value corresponds to replacing some of the contexts with a single joint model, while the "collapsing" p-value is associated to pruning all the sub-contexts of a context. P-values associated to specific models correspond to reducing the memory of the corresponding model, that is discarding the dependency 
of the conditional probability towards the oldest covariates. 

To explore the contexts in a programmatically way, one should rely on the `contexts()` function. COVLMC contexts have additional characteristics compared to VLMC and context trees.
In particular, the `contexts()` function can report the model associated to each context, either
by its parameters only:
```{r}
contexts(elec_model, model = "coef")
```
or using the models themselves:
```{r}
contexts(elec_model, model = "full")
```

See `contexts.covlmc()` for details.
